---
title: "Reproducible Research: Peer Assessment 1"
author: "Jack Palmer"
date: "August 6, 2016"
output: 
  html_document: 
    keep_md: yes
---

```{r load packages}
library(ggplot2)
library(lattice)
options(scipen = 999)
```

Load the data

```{r load data}
temp <- tempfile()
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",temp)
data <- read.csv(unz(temp, "activity.csv"))
unlink(temp)
data$date <- as.Date(data$date)
```

## What is mean total number of steps taken per day?

Aggregate the steps by day to calculate the total number of steps taken by day.

```{r aggregate steps by day}
steps_by_day <- aggregate(data$steps, by = list(data$date), sum)
colnames(steps_by_day) <- c("date","steps")
```

We then make a histogram of the total number of steps taken each day.

```{r histogram of steps by day}
m <- ggplot(data=steps_by_day, aes(x=steps)) 
m + geom_histogram(binwidth = 1000) + xlab("Steps") + ylab("Frequency")
```

Calculate and report the mean and median of the total number of steps taken per day.

```{r mean and median steps by day}
mean_steps_by_day <- mean(steps_by_day$steps, na.rm = T)
median_steps_by_day <- median(steps_by_day$steps, na.rm = T)
```

The mean number of steps taken each day is `r mean_steps_by_day`. The median number of steps taken each day is `r median_steps_by_day`.

## What is the average daily activity pattern?

Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r time series plot of average steps taken by interval}
mean_steps_by_interval <- aggregate(data$steps, by = list(data$interval), mean, na.rm = T)
colnames(mean_steps_by_interval) <- c("interval","steps")
plot(mean_steps_by_interval$interval,mean_steps_by_interval$steps, type="l", xlab= "Time Interval", ylab= "Average Number of Steps Taken")
```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r average steps by interval}
mean_steps_by_interval <- mean_steps_by_interval[order(-mean_steps_by_interval$steps), ]
top_interval <- mean_steps_by_interval$interval[1]
```

The interval `r top_interval` contains the maximum number of steps on average across all days.

## Imputing missing values

Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r identify missing values}
missing_values <- sum(is.na(data$steps))
```

There are `r missing_values` missing values in the dataset.

Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r clean data set and add average steps for interval}
data_cleaned <- merge(data,mean_steps_by_interval,by.x = c("interval"),by.y = c("interval"),all.x = T)
data_cleaned$steps <- ifelse(!is.na(data_cleaned$steps.x),data_cleaned$steps.x,data_cleaned$steps.y)
```

Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

Aggregate the steps by day to calculate the total number of steps taken by day.

```{r steps by day with cleaned data}
steps_by_day_cleaned <- aggregate(data_cleaned$steps, by = list(data_cleaned$date), sum)
colnames(steps_by_day_cleaned) <- c("date","steps")
```

Make histogram

```{r histogram of steps by day with cleaned data}
m <- ggplot(data=steps_by_day_cleaned, aes(x=steps)) 
m + geom_histogram(binwidth = 1000) + xlab("Steps") + ylab("Frequency")
```

Calculate the difference in total number of steps

```{r calculate step difference}
step_difference <- sum(data_cleaned$steps) - sum(data$steps, na.rm = T) 
```

Imputing the missing data results in an additional `r step_difference` steps.

Calculate and report the mean and median of the total number of steps taken per day for the cleaned data set.

```{r mean and median steps by day for cleaned data}
mean_steps_by_day_cleaned <- mean(steps_by_day_cleaned$steps, na.rm = T)
median_steps_by_day_cleaned <- median(steps_by_day_cleaned$steps, na.rm = T)
mean_steps_differential <- mean_steps_by_day - mean_steps_by_day_cleaned
median_steps_differential <- median_steps_by_day - median_steps_by_day_cleaned
```

The difference in the mean number of steps is `r mean_steps_differential`. The difference in the median number of steps is `r median_steps_differential`.

## Are there differences in activity patterns between weekdays and weekends?

Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r assign dow factor}
data$dow <- ifelse(weekdays(data$date) == "Saturday" | weekdays(data$date) == "Sunday", "Weekend", "Weekday")
```

Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 

```{r time series panel plot}
mean_steps_by_interval_dow <- aggregate(data$steps, by = list(data$interval,data$dow), mean, na.rm = T)
colnames(mean_steps_by_interval_dow) <- c("interval","dow","steps")
xyplot(steps~interval | factor(dow), data=mean_steps_by_interval_dow,type = "l", xlab = "Interval", ylab = "Number of steps",layout=c(1,2))
```